---
title: "10. Probability distributions"
output: pdf_document
urlcolor: blue
---

These notes consider the Chapter 10 of the [handbook](https://emotion.utu.fi/tilastotiede/) on various probability distributions.

# Discrete distributions

For random variables with countable number of possible values.

## Binomial distribution

See page 200 for proper definition.

```{=latex}
Useful when a random variable $X$ has exactly two exclusive possible outcomes (e.g. success/fail) with known probabilities $p\in[0,1]$ (success) and $q=1-p$ (fail). For $n\in\mathbb{N}$ trials with $k=0,1,2,\dots ,n$ successes

\begin{displaymath}
X\sim Bin(n,p)
\end{displaymath}

Probability Mass Function (PMF)

\begin{displaymath}
\mathrm{P}(X=k)=\binom{n}{k}p^kq^{n-k}
\end{displaymath}

Cumulative Distribution Function (CDF)

\begin{displaymath}
\mathrm{F}(k)=\mathrm{P}(X\leqslant k)=\sum_{i=0}^k\binom{n}{i}p^iq^{n-i}
\end{displaymath}

In the following code examples, equal probabilities are assumed, $n=$ \textbf{size}, $p=$ \textbf{prob} $=0.5$.
```

```{r out.width = "75%", fig.align = "center"}
# Sequence for visualization
binomial_seq <- seq(0, 25, by = 1)

# Functions
binomial_pmf <- dbinom(x = binomial_seq, size = 25, prob = 0.5)
binomial_cdf <- pbinom(q = binomial_seq, size = 25, prob = 0.5)

# Plot
par(mfrow = c(1, 2))
plot(binomial_seq, binomial_pmf, ann = FALSE)
plot(binomial_seq, binomial_cdf, type = "S", ann = FALSE)
```

```{r out.width = "75%", fig.align = "center"}
# Probability for exactly 3 successes out of 10 trials
dbinom(x = 3, size = 10, prob = 0.5)

# Probability for up to 3 successes out of 10 trials
pbinom(q = 3, size = 10, prob = 0.5)

# Simulate 10 times how many successes there is using random numbers
rbinom(n = 10, size = 10, prob = 0.5)

# With large enough n, expected value (np = 5) should become visible
rbinom500 <- rbinom(n = 500, size = 10, prob = 0.5)
hist(rbinom500, xlab = "Number of successes", ylab = "Frequency", main = NULL)
summary(rbinom500)
```

## Poisson distribution

See page 204.

```{=latex}
Useful when estimating amounts in random processes where the expected value $(\lambda)$ is known

\begin{displaymath}
X\sim Poi(\lambda)
\end{displaymath}

PMF

\begin{displaymath}
\mathrm{P}(X=k)=\frac{\lambda^ke^{-\lambda}}{k!} \;,\; k=0,1,2,\dots ,n
\end{displaymath}

Note that the Poisson distribution can be used to approximate the binomial distribution when $n$ is large and $p$ is small. In this case, $\lambda=np$ and

\begin{displaymath}
\mathrm{P}(X=k)=\frac{(np)^ke^{-np}}{k!}
\end{displaymath}
```

```{r out.width = "75%", fig.align = "center"}
# Sequence for visualization
poisson_seq <- seq(0, 50, by = 1)

# Functions
poisson_pmf <- dpois(x = poisson_seq, lambda = 20)
poisson_cdf <- ppois(q = poisson_seq, lambda = 20)

# Plot
par(mfrow = c(1, 2))
plot(poisson_seq, poisson_pmf, ann = FALSE)
plot(poisson_seq, poisson_cdf, type = "S", ann = FALSE)
```
```{r out.width = "75%", fig.align = "center"}
# A bridge is crossed by 12 people per minute, on average.
# What is the probability that exactly 16 people crosses it in a minute?
dpois(x = 16, lambda = 12)

# A bridge is crossed by 12 people per minute, on average.
# What is the probability that up to 16 people crosses it in a minute?
ppois(q = 16, lambda = 12)

# Simulate amount of people per minute 10 times
rpois(n = 10, lambda = 12)

# By repeating the observation once every day for a year,
# the expected value (lambda = 12) should become visible
poisson_pedestrians <- rpois(n = 365, lambda = 12)
hist(poisson_pedestrians, breaks = 20,
     xlab = "Number of pedestrians", ylab = "Frequency", main = NULL)

summary(poisson_pedestrians)
```

# Continuous distributions

For random variables with infinite number of possible values.

## Generic definitions

See page 208.

```{=latex}
The probability that an event happens between an interval $[a,b]$ can be calculated from the Probability Density Function (PDF) $f(x)$ as an integral

\begin{displaymath}
\mathrm{P}(a\leqslant X\leqslant b)=\int_a^bf(x) \;\mathrm{d}x
\end{displaymath}

Note that the point probabilities for continuous random variables are intrinsically zero

\begin{displaymath}
\int_a^af(x) \;\mathrm{d}x = 0
\end{displaymath}

CDF

\begin{displaymath}
\mathrm{P}(X\leqslant t)=\int_{-\infty}^tf(x) \;\mathrm{d}x
\end{displaymath}
```

## Normal distribution

See page 210.

```{=latex}
Many phenomena in nature follow the normal distribution. Additionally, so called central limit theorem states that the averages of idenpendet random samples drawn from a population approximately follow the normal distribution - even if the population is better desrcibed by some other distribution!

Normal distribution is defined by expected value (and median and mode due to symmetry) $\mu$ and (standard) deviation $\sigma$.

\begin{displaymath}
X\sim N(\mu,\sigma^2)
\end{displaymath}

PDF

\begin{displaymath}
f(x)=\frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}
\end{displaymath}

In the following code examples, $\mu=$ \textbf{mean}, $\sigma=$ \textbf{sd}.
```

```{r out.width = "75%", fig.align = "center"}
# Sequence for visualization
normal_seq <- seq(-5, 5, by = 0.01)

# Functions
normal_pmf <- dnorm(x = normal_seq, mean = 0, sd = 1)
normal_cdf <- pnorm(q = normal_seq, mean = 0, sd = 1)

# Plot
par(mfrow = c(1, 2))
plot(normal_seq, normal_pmf, type = "l", lwd = 3, ann = FALSE)
plot(normal_seq, normal_cdf, type = "l", lwd = 3, ann = FALSE)
```

```{r out.width = "75%", fig.align = "center"}
# The average length of a zebrafish is 22 mm with a standard diviation of 1 mm.
# How likely is a fish 23 mm long?
dnorm(x = 23, mean = 22, sd = 1)

# What fraction of the fish are up to 23 mm long?
pnorm(q = 23, mean = 22, sd = 1)

# Simulate length of five random fish
rnorm(n = 5, mean = 22, sd = 1)

# With large enough n, expected value should become visible
normal_zebrafish <- rnorm(n = 400, mean = 22, sd = 1)
hist(normal_zebrafish, breaks = 20,
     xlab = "Fish length", ylab = "Frequency", main = NULL)
summary(normal_zebrafish)
```

## Student's t-distribution

See page 215.

```{=latex}
Similar to the normal distribution, Student's $t$ distribution can be used to estimate the expected value of a population using large enough amout of sample averages.

The Student's $t$ distribution is defined by just one parameter, namely the degrees of freedom (DoF) $\nu$. Compared to the normal distribution, the advantage here is that prior knowledge of parameters $\mu$ and $\sigma$ is not needed - which oftentimes is the case. Additionally, with a large $\nu$, the Student's $t$ distribution approaches the standard normal distribution N(0,1). 

PDF is defined using the Gamma function $\Gamma(n)=(n-1)!$.

\begin{displaymath}
f(t)=\frac{\Gamma(\frac{\nu+1}{2})}{\sqrt{\pi\nu}\Gamma(\frac{\nu}{2})}\left(1+\frac{t^2}{\nu}\right)^{-(\nu+1)/2}
\end{displaymath}

In the following code examples, $\nu=$ \textbf{df}.
```

```{r out.width = "75%", fig.align = "center"}
# Sequence for visualization
students_t_seq <- seq(-5, 5, by = 0.01)

# Functions
students_t_pmf <- dt(x = students_t_seq, df = 10)
students_t_cdf <- pt(q = students_t_seq, df = 10)

# Plot
par(mfrow = c(1, 2))
plot(normal_seq, students_t_pmf, type = "l", lwd = 3, ann = FALSE)
plot(normal_seq, students_t_cdf, type = "l", lwd = 3, ann = FALSE)
```

```{r}
# Same as in the earlier examples
dt(x = 2, df = 20)
pt(q = 2, df = 20)
rt(n = 5, df = 20)

# The mean should be around 0 regardless of DoF
student_values <- rt(n = 200, df = 5)
hist(student_values, breaks = 20,
     xlab = "t-values", ylab = "Frequency", main = NULL)
summary(student_values)
```
